A Linear Solution of the Linearity Challenge

### 2025-12-19 paper/related.tex
related

The rightmost columns of \Cref{tab:comparison} report the \emph{size} (in
kilobytes) of the formalizations as rough (and possibly questionable) estimates
of their complexity. Instead of counting line numbers, which may depend on code
indentation styles and syntactical constraints of the proof assistant being
used, we have counted the number of characters after comments have been removed
and spaces have been squeezed.\footnote{Sequences of two or more consecutive
space-like characters are collapsed into a single space. The squeezing is
obtained by running the command \texttt{tr -s [:space:] file} on Unix-like
systems.}
%
The reported sizes account for the source code of the formalization excluding
examples and any ``safety'' result, if present. We have noticed that the
interpretation of ``safety'' varies widely among the formalizations and, except
for our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24} when available.
%
Some formalization define \emph{libraries} which could be reused in different
contexts. In these cases, the size of the library is reported separately next to
the size of the part of the development that uses it.

In general, it is difficult to draw firm conclusions on the effectiveness of the
various approaches because the formalizations differ substantially in terms of
reference calculus, feature coverage, proof assistant,
process/expression/channel representation, handling of linearity. Also, most of
them predate the linearity challenge~\cite{CarboneEtAl24}, hence they are not
necessarily targeted towards that particular problem. However, some of
them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
are explicitly motivated by the complexity arising from the linear usage of
resources/channels and from context splitting in particular.

we can make some inference based on the available data.

paper di Marco a ECOOP (multiparty)

>>>>

The rightmost columns of \Cref{tab:comparison} report the \emph{size} (in
kilobytes) of the formalizations as rough (and possibly questionable) estimates
of their complexity. Instead of counting line numbers, which may depend on code
indentation styles and syntactical constraints of the proof assistant being
used, we have counted the number of characters after comments have been removed
and spaces have been squeezed.\footnote{Sequences of two or more consecutive
space-like characters are collapsed into a single space. The squeezing is
obtained by running the command \texttt{tr -s [:space:] file} on Unix-like
systems.}
%
The reported sizes account for the source code of the formalization excluding
examples and any safety result, if present. We have noticed that the
interpretation of ``safety'' varies widely among the formalizations and, except
for our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24} when available.
%
Some formalization define \emph{libraries} which could be reused in different
contexts. In these cases, the size of the library is reported separately next to
the size of the part of the development that uses it.

In general, it is difficult to draw firm conclusions on the effectiveness of the
various approaches in addressing the linearity challenge because the
formalizations differ substantially in terms of reference calculus, feature
coverage, proof assistant, process/expression/channel representation, handling
of linearity. Also, most of them predate the publication of the linearity
challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted towards
that particular problem. However, some of
them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
are explicitly motivated by the complexity arising from the linear usage of
resources/channels and from context splitting in particular.

Based on the available data, we can make the following observations.
%
Context splitting, which is very well represented, does not seem to be a good
indicator of the complexity of the formalization, since it spans the whole range
of formalization sizes. In particular, our formalization is based on context
splitting and covers full \CP~\cite{Wadler14}, yet it is by far the smallest
one.

The largest formalizations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are also
the ones that adopt a locally nameless representation of channels. In this
formalizations channels are represented in two different ways, depending on
whether they are free or bound. This necessarily entails some duplication of
effort as well as some transformation machinery between the two representations
that contribute to the complexity of the formalization.
%
Other channel representations are not strong complexity indicators. Note that
the adoption of co-de Bruijn syntax implies the use of context splitting, hence
the two aspects are not completely independent.

There is no strong evidence that the intrinsically typed representation of
processes reduces the size of the formalization. As observed in
\Cref{sec:process-agda}, this choice may help reducing the overall number of
datatypes to be defined, but the definitions are necessarily more involved. In
the end, it seems plausible to speculate that the effort for representing
processes, types and typing rules is not substantially different overall.

Finally, it seems that the use of (explicit) continuations helps taming the
complexity of the formalization more than anything else. Indeed, the six
smallest formalizations (excluding the deadlock freedom results) -- with an
average size of around 62kb -- are all based on continuations, whether they are
explicit (\LinearPI, \SCP, \LCP) or ``virtual'' (\GV), while the remaining ones
are 263kb on average. This observation is in agreement with the observations we
made earlier in this paper concerning the difference between \emph{updating}
bindings versus \emph{removing and adding} bindings in a typing context. The
adoption of continuations enables a clean linear management of channels and
avoids the need for strong updates on the typing context.

% It is also interesting to note that
% whereby \SCP, which the authors~\citet{SanoKavanaghPientka23} do not relate
% directly to the linear $\pi$-calculus, relies on explicit continuations to tame
% the complexity of resource management. Our comparison suggests that this aspect,
% more than the structural treatment of contexts,

% \citet{SanoKavanaghPientka23} do not relate it directly to the linear
% $\pi$-calculus.

% più che il trattamento strutturale dei contesti (ed evitare lo splitting) è
% l'uso delle continuazione a fare la differenza

In this work we have focused on models and theories of \emph{binary sessions},
but there are also formalizations of \emph{multiparty session types}, notably
those of \citet{JacobsBalzerKrebbers22a} and \citet{TiroreBengstonCarbone25},
which are notoriously more complex than those of binary session types. Just to
give an example, the formalization of \citet{TiroreBengstonCarbone25} amounts to
more than 1Mb of Coq code.
